Psych 254 W15 PS #2

### 2015-02-16 analyses/ps2.Rmd
Some progress on ps2
Finished ggplot assignment

---
title: 'Psych 254 W15 PS #2'
author: "Mike Frank"
date: "February 4, 2015"
output: html_document
---

@@

```{r}
install.packages("ggplot2")
library(ggplot2)
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

sem <- function(x) {sd(x) / sqrt(length(x))}
ci95 <- function(x) {sem(x) * 1.96}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

@@

```{r} 
```

@@

```{r}
```

@@

Sklar et al. show a plot of a "facilitation effect" - the time to respond to incongruent primes minus the time to respond to congruent primes. They then show plot this difference score for the subtraction condition and for the two presentation times they tested. Try to reproduce this analysis.

@@

```{r}
```

@@

```{r}
```

@@

@@

```{r}
```

@@

```{r}
```

@@

```{r}
```

>>>>

---
title: 'Psych 254 W15 PS #2'
author: "Robert Hawkins"
date: "February 16, 2015"
output: html_document
---

@@

```{r}
library(ggplot2)
setwd("~/Repos/psych254/analyses/")
```

@@

```{r}
qplot(carat, price, data = diamonds)
```

@@

```{r}
qplot(carat, price, col=clarity, shape=cut, data = diamonds)
```

@@

```{r}
qplot(carat, price, facets = clarity ~ cut, data = diamonds)
```

@@

```{r}
qplot(carat, price, col = clarity, facets = ~ cut, data = diamonds)
```

@@

```{r}
qplot(price, geom="histogram", data = diamonds)
```

**We see that most diamonds are between \$0 and \$1000 (with a mode around \$725), but that there is a very long tail, dragging the mean up to \$4000**

@@

```{r}
qplot(price, geom="histogram", facets =clarity ~ cut , data = diamonds)
```

**Unsurprisingly, the tail gets longer as the clarity gets bettter, although the VVS2 cut appears to have fairly long tails regardless of cut**

@@

```{r}
qplot(price, geom="histogram", facets =clarity ~ cut , data = diamonds) + theme_bw()
```

@@

```{r}
d <- (ggplot(diamonds, aes(x = carat, y=price))
      + geom_point(aes(colour = carat)))
d
```

@@

```{r}
d <- (ggplot(diamonds, aes(x = price))
      + geom_histogram())
d
```

@@

sem <- function(x) {sd(x, na.rm = T) / sqrt(length(x))}
ci95 <- function(x) {sem(x) * 1.96}
```

@@

```{r}
d.a.long = gather(d.a, subid, RT, starts_with("X")) %>% 
  mutate(subid = as.numeric(substr(subid, 2,3))) 
d.b.long = gather(d.b, subid, RT, starts_with("X")) %>% 
  mutate(subid = as.numeric(substr(subid, 2,3)))
```

@@

```{r}
d.data <- bind_rows(d.a.long, d.b.long)
cat(length(d.a.long$subid), length(d.b.long$subid), length(d.data$subid))
```

@@

```{r}
d <- right_join(d.data, subinfo, by = c('subid'))
```

@@

```{r}
library(ggplot2)
qplot(RT, data = d)
```

**We see that the data is a bit skewed to the left but otherwise passably normal, centered at about 600ms. We also note a few outliers toward the top and bottom.**

@@

```{r}
qplot(RT, subid, geom = 'point', data = d)
```

**First, we note that if we plot individual RTs for each participant, they seem to fall in bins rather than continuously**

```{r}
unique_RTs = sort(unique(subset(d, RT <= 1000 & RT >= 500)$RT))
qplot(1:length(unique_RTs), unique_RTs)
cat("sample rate is approximately", (length(unique_RTs)/3)*2, "samples per second")
```

**To make this observation precise, we look at the interval between 500ms and 1000ms where the most data was collected. We pool together all RTs from all participants in all conditions, record the unique values, and sort them from lowest to highest. This reveals a clear 'binning' pattern, with rounding error or tolerance leading to three values in each bin (1ms above and below the central value). There were 14 bins (with 3 unique values each) between 500ms and 100ms, which we extrapolate to a sampling rate of 28 samples per second.**

@@

```{r} 
qplot(objective.test, subjective.test,  data = d)
log_fit = glm(subjective.test ~ objective.test, family = binomial, data = d)
summary(log_fit)
xr = seq(.3, 1, .05)    # possible values of Responsible
yc = predict(log_fit, data.frame(objective.test = xr), type = 'response')
logit.data <- data.frame(xr,yc)
ggplot(d, aes(x=objective.test, y=subjective.test)) +
    geom_rect(xmin = -0.1, ymin = -0.1, xmax = .6, ymax = .1, fill = 'red') +
  geom_point() +
  geom_line(data = logit.data, aes(x = xr, y = yc)) 
```

**We ran a logistic regression predicting the probability of subjectively seeing the primes from performance on an objective 2AFC task about their parity. We find that people who reported subjectively seeing the primes were also more likely to objectively do well on the 2AFC test about primes. $b = 9.36, t(6467) = 37.1, p < 0.001$ **

@@

```{r}
ds <- d %>% 
  filter(subjective.test == 0 & objective.test < .6)
cat("excluded", length(unique(d$subid)) - length(unique(ds$subid)), "out of ",
    length(unique(d$subid)))
```

@@

Sklar et al. show a plot of a "facilitation effect" - the time to respond to incongruent primes minus the time to respond to congruent primes. They then plot this difference score for the subtraction condition and for the two presentation times they tested. Try to reproduce this analysis.

@@

```{r}
RT_congruent <- ds %>% 
  filter(congruent == "yes") %>% 
  group_by(subid, operand, presentation.time) %>% 
  summarise(congruent_mean = mean(RT, na.rm = T))
RT_incongruent <- ds %>% 
  filter(congruent == "no") %>%
  group_by(subid, operand, presentation.time) %>% 
  summarise(incongruent_mean = mean(RT, na.rm = T))
RT_sum <- right_join(RT_congruent, RT_incongruent) %>%
  mutate(diff_mean = incongruent_mean - congruent_mean) %>%
  group_by(presentation.time, operand) %>%
  mutate(error = sem(diff_mean)) %>%
  select(subid, presentation.time, operand, error, diff_mean) %>%
  summarize(error = mean(error), 
            overall_mean = mean(diff_mean, na.rm = T))
head(RT_sum)
```

@@

```{r}
d <- (ggplot(RT_sum, aes(x = operand, y = overall_mean, 
                         fill = presentation.time))
      + geom_bar(stat = "identity", position = "dodge")
      + geom_errorbar(aes(ymax = overall_mean + error, 
                          ymin=overall_mean - error),
                      position = position_dodge(width=0.9),
                      width = 0.25)
      + ylab("Facilitation (ms)")
      + scale_fill_discrete(name="Presentation Duration"))
d
```

@@

**The "subtraction" part of our graph is numerically identical to the one presented by Sklar et al, and our interpretation of this result is the same: there is a greater facilitation effect for shorter display durations than longer display durations in the subtraction condition. We also note some differences in presentation:**

1) **Note that our error bars appear much larger. Upon further inspection, this can be attributed to what they mean by "SEM." Typically, this means the "standard error of the mean" (i.e. $\bar{x} \pm 1 SE$, which is what we plotted), but they appear to have plotted the interval containing exactly one standard error.**

2) **The tick mark labels on their $y$-axis show three (redundant) decimal places, making the numbers look bigger.**

3) **They decided not to plot results for the 'addition' condition rather than including it in the plot.**

**Overall, following remarks made by Andrew Gelman and others on experimenter degrees of freedom, we must express some skepticism about the interpretation offered by the authors. If an effect was found in the addition but not the subtraction condition, or in both conditions, would other justifications have been offered? If the effect were reversed such that longer presentation durations led to greater facilitation, would the authors be able to justify this as well? What if only one or the other of the manipulation checks were used, or the threshold in the objective test were moved down? Since the actual size of the effect is quite small (as seen from the error bars), further work seems to be needed in justifying these decisions.**

@@

**Didn't get around to this**

@@

**If arithmetic could be done 'non-consciously', we wouldn't necessarily expect presentation time or operand to matter. We would just want to know whether or not there was a significant facilitation effect across the board.**

```{r}
RT_sum2 <- right_join(RT_congruent, RT_incongruent) %>%
  mutate(diff_mean = incongruent_mean - congruent_mean) %>%
  group_by(subid) %>%
  summarize(overall_mean = mean(diff_mean, na.rm = T)) %>%
  mutate(error = sem(overall_mean))

pop_mean = mean(RT_sum2$overall_mean)
error = mean(RT_sum2$error)
d <- (ggplot(RT_sum2, aes(x = overall_mean))
      + geom_rect(aes(xmin=pop_mean - 2*error, 
                xmax=pop_mean + 2*error, 
                ymin=0, ymax=Inf), 
            fill = 'red', alpha = 0.1)
      + geom_histogram()
      + geom_vline(xintercept = pop_mean)
)
d
```

**We plot a histogram of each participant's average facilitation effect across conditions, and show the approximate 95% confidence interval ($\bar{x} \pm 2SE$) for the facilitation effect estimate. It is not clear from the graph whether it is significant -- the confidence interval includes 0 -- but it is certainly trending in the direction of the claim for non-conscious arithmetic**

@@

**We simply run a t-test to check whether the distribution of individual facilitation effects is different from 0**

```{r}
t.test(RT_sum2$overall_mean)
```

**We find that the facilitation effect is marginally greater than 0 ($m = 5.37, t(16) = 1.84, p = 0.084$), but not significant. Sklar et al.'s finding are based on various ways of slicing these data into subsets which are significantly different from other subsets, but the fact that there is the overall facilitation effect is small (5ms) and only marginally significant shows that whatever non-conscious processes contribute to the way we do arithmetic are sensitive to a variety of other factors (including at least presentation time and operand). This same conclusion is given by the authors at the end of their results: "What the specific conditions are under which addition and subtraction equations are nonconsciously solved turns out to be a complex issue that must be left for future investigations."**
